Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2017-10-29 output/breed_paper.Rnw
Commit before inverting the sections in results.

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
better understand what improvements have been made in the genetic
yield potential obtained through plant breeding. Specifically, the
hypothesis was tested that the yield of a rice cultivar declines over
time after its introduction. This would explain how
side-by-side comparisons show increased yield while no substantive
increases have been seen in farmers' fields. If there is yield
erosion, side-by-side comparisons will show ``false'' yield increases due to the
yield of the older cultivar decreasing, not due to the increased
genetic yield potential of the newer cultivar. An apparent increase in
yield in side-by-side comparisons between old and new rice varieties
was observed when the yield erosion was not considered. However, this
apparent increase in yield could be explained by the differences in
the time when these cultivars were selected. Yield
eroded at an estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\%
credible interval -4.4 to -53.3) after initial selection. Once this
effect was considered, the yield advantage of newer cultivars
over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
credible interval -36.1 to 31.5). Thus, without continuous plant improvement and
deployment of new cultivars, yield will not only remain stagnant but
will decline over time. To increasing the genetic yield potential of
rice in the face of this phenomenon, higher yield targets are needed
to counter-act the decrease over time in the current cultivars.

@@

% \linenumbers

@@

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area \citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However production systems are experiencing plateaus in grain
yield \citep{Grassini2013, Peng2010}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications. Therefore, it is of critical importance to better
understand why the rate of increase in grain yields has declined or
leveled off in intensified production systems.

California (CA) rice production represents one such production
system. Production in CA is predominately focused on premium quality
medium grain \textit{japonica} cultivars (e.g., CalRose rice), and
rice from CA is recognized globally for its quality. Rice yields in CA
increased rapidly during the period from 1920 -- 1990
(Fig. \ref{fig:state_yield}), in part due to improvements in rice
genetics (i.e., the ``Green Revolution''). Most cultivars in use in CA
are developed by the California Cooperative Rice Research Foundation
(CCRRF; a collaboration between the University of California, the
USDA-Agricultural Research Service, and farmer-funded
research). Although grain quality has been maintained or even improved, rate of yield
increase has slowed since the 1990s (Fig. \ref{fig:state_yield})
despite continuous crop improvement \citep{ccrrf}. For the past 30
years, each newly developed cultivar has demonstrated a yield
advantage over the previous cultivars in the Statewide Variety Trials
(an annual multi-site variety trial program that compares experimental
entries side-by-side with ``check'' cultivars). However, in a seeming
contradiction these same reports show stagnant yields across release
announcements (i.e., reports accompanying the official public release)
from the previous years. For example, when cultivar M-202 was released
in 1985, it was reported to yield an average of 11 Mg ha\tsups{-1}
\citep[14\% moisture content (MC);][]{m202}. In 2015, the reported
average yield of the newest released cultivar M-209 was 10.8 Mg
ha\tsups{-1} \citep]14\% MC;][]{m209}. In other rice systems, a
decrease in a cultivar's yield performance over time has been observed
\citep{Peng1998, Peng2010, Peng2000, Datta1995}, often termed yield
erosion \citep{Peng1998}. It has been speculated that this yield
erosion is due to the inability of a cultivar to adapt to changing
biotic and abiotic conditions. To better understand how plant
improvement and yield erosion are contributing to the improvement of
rice yield in farmers' fields, we teseted the hypotheses: either (1)
the genetic yield advantage increased over the last 30+ years, or (2)
the genetic yield advantage has not increased over the last 30+
because the erosion of yield over time creates an illusion of improved
yields.

@@

The CCRRF Rice Experiment Station (RES) is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice cultivars adapted to CA since
1969. The climate at the RES is Mediterranean, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} to \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% soil organic matter
\citep{ssurgo} in the top 15cm. As the central location for the CCRRF
breeding program, the RES is researcher-managed (as opposed to
farmer-managed trials) and plants the majority of experimental plots for the
Statewide Variety Trials (approximately 3 to 4 times the number of
experimental plots compared to any other location). To avoid potential
complications due to site, climate, and management differences at the
other 6 to 8 farmer-managed trials per year we focused our
investigation on the RES.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. These cultivars are typcically
evaluated in the Statewide Variety Trials for 3 to 5 years prior to
their official release. For these cultivars planted in breeding trials
at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
3--4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars,
including every publicly released medium grain cultivar developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release was
calculated.

Experimental plots at the RES are managed similar to University of California
Cooperative Extension following prescribed best management
practice, including land preparation, fertility management, and pest
and disease control \citep{ucce}. For planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\tsups{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

To test each hypothesis, the influence of breeding and yield erosion
were quantified on yield over time. Yield was modeled as a response to
the number of years since the cultivar was officially released and the
year of release using a mixed-effects linear regression. To account
for similarities between years and between cultivars, year and
cultivar were included in the model as random effects.

@@

Data were processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the `rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package \citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval \citep{rstanarm}. Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

@@

Comparing the cultivars' yield during each's initial time (average 3 years) in the
experimental trials confirmed that genetic yield improvements have
been consistant over the 30 year study period
(Fig. \ref{fig:first_yrs}). As estimated by the model including both
yield erosion and yield improvement over time, there is
no conclusive evidence that the yields of medium grain rice cultivars
selected and released by CCRRF's breeding program have, on average,
increased over the last 30 years (Table \ref{tbl:coef}). The full
model which corrected for both the year of release and yield erosion
was estimated to have better out-of-sample predictive accuracy
compared to reduced models containing only one predictor (See
supplemental material for further details). The
estimated change in grain yield per year in the CCRRF's breeding
trials at the RES was estimated to be a negligible
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:first_yrs}).

@@

plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
     yaxt = "n",
     ylim = c(8, 12.5),
     xaxt = "n",
     xlim = c(1980,2020))
axis(1, at = seq(1980, 2020, by = 10))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000,
       lty = 2)
with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
                         y1 = mean - se))
text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
     labels = gsub("M", "M-", row.names(plot_data2)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
mtext("Adjusting for yield erosion", adj = 0)
polygon(x = c(yrs, rev(yrs)),
        y = c(ints2[1,],rev(ints2[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since release",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8.0, 12.5))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
       b = yrs_coef/1000)
polygon(x = c(yrs_in, rev(yrs_in)),
        y = c(ints[1,],rev(ints[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))
text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))

@@

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other advancements due to plant
improvement, including increased quality \citep{mckenzie1993} and
milling yield \citep{mckenzie1994}, have been made over the study
period and are important. Also, crop duration has been reduced. For
example, during this study period, M-202 and M-206 have been the
predominate cultivars grown in CA, and while their yield potentials
were similar at the time of release (Fig. 5), M-206 matures about 6
days earlier than M-202 (Table 2). Shorter duration is very important
in CA due to an already short growing season and water limitations
(shorter duration cultivars would presumably require less irrigation
water).

>>>>

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
quantify improvements made in the genetic yield potential obtained
through plant breeding. California rice systems are among the highest
in the world and close to maximum yield potential. Specifically, the
hypothesis was tested that if rice cultivar yields decline over time
(termed yield erosion) then apparent yield increases in side-by-side
yield comparison tests will not reflect increases in yield
potential. This hypothesis was tested using experimental yield data
over 30+ years at the California Cooperative Rice Research Foundation
Rice Experiment Station. Results show an apparent increase in yield in
side-by-side comparisons between old and new rice varieties was
observed when the yield erosion was not considered. However, the
yields of older varieties eroded at an estimated rate of 29.3 kg
ha\tsups{-1} year\tsups{-1} (90\% credible interval -4.4 to -53.3)
after initial selection. Once this effect was considered, the yield
advantage of newer cultivars over old was uncertain (-3.3 kg
ha\tsups{-1} year\tsups{-1}, 90\% credible interval -36.1 to
31.5). These results highlight the importance of continuous crop
improvement and deployment of new cultivars as yield will decline over
time without it. To increase the genetic yield potential of rice in
the face of this phenomenon, higher yield targets are needed to
counter-act the decrease over time in the current cultivars.

@@

%\linenumbers

@@

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area \citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However, many production systems are experiencing plateaus in grain
yield \citep{Grassini2013, Peng2010}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications~\citep{Foley2011a}. Therefore, it is of critical importance to better
understand why the rate of increase in grain yields has declined or
leveled off in intensified production systems.

California (CA) rice represents one such production system. Rice is
grown primarily in the Sacramento Valley which is characterized by
having a Mediterranean climate, with long days, dry growing season,
and relatively free of pests and diseases, which lead to some of the
highest yields in the world~\citep{FAO2016}. Production in CA is
predominately focused on premium quality medium grain
\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
recognized globally for its quality
(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
cultivars in use in CA are developed by the California Cooperative
Rice Research Foundation (CCRRF; a collaboration between the
University of California, the USDA-Agricultural Research Service, and
farmer-funded research). In part due to improvements in rice genetics,
rice yields in CA increased rapidly during the period from 1920 --
1990; however, since the 1990s, the rate of yield increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop
improvement. Based on a yield-gap analysis, on-farm yields in the
major rice growing region of CA is 73 to 76\% of maximum yield
potential~\citep{espe16a}. In highly intensive systems such as this,
\citet{grassini2011} have shown that farmers are capable of attaining
85\% of the maximum yield potential. Thus, given that farmers are near
the attainable yield potential, increases in yield are likely to be
slow and \citet{espe16a} reported that indeed this was the case
with yields on average increasing by about 50 to 62 kg ha-1 between
1999 and 2014.

Certainly one question in looking at this situation is, “do new
varieties have increased yield potential”? Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding varieties with increased yield potential. However, when the
popular cultivar M-202 was
released in 1985, it was reported to yield an average of 11 Mg
ha\tsups{-1} \citep[14\% moisture content (MC);][]{m202}. In 2015, the
reported average yield of the newest released cultivar M-209 was
10.8 Mg ha\tsups{-1} \citep[14\% MC;][]{m209}. However, in annual
breeding reports and varietal release announcements, based on yield
performance in side-by-side variety trails, data show that new
varieties usually out yield the older varieties. In other rice
systems, a decrease in a cultivar's yield performance over time has
been observed \citep{Peng1998, Peng2010, Peng2000, Datta1995}, often
termed yield erosion \citep{Peng1998}. It has been speculated that
this yield erosion is due to the inability of a cultivar to adapt to
changing biotic and abiotic conditions~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}. Yield erosion may explain the apparent
contradiction above, in which side-by-side yield comparisons show
yield improvements in varieties while overall there may be little
change in yield potential. To address this issue we teseted the
hypothesis that if rice cultivar yields erode over time then apparent
yield increases in side-by-side yield comparison tests will not
reflect true yield increases.

@@

The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
(39.4648, -121.7342), and has been the central location for CCRRF's
efforts to develop improved rice cultivars adapted to CA since
1969. The California Statewide Variety Trails evaluates current and
promising varieties at the RES and six to eight on-farm trials around
the state each year. To avoid potential complications due to site,
climate, and management differences at the farmer-managed trials per
year, we focused our investigation on the RES. The RES is
researcher-managed and plants the majority of the experimental plots
for the statewide program (approximately 3 to 4 times the number of
experimental plots compared to other on-farm locations). The climate
at the RES is Mediterranean, characterized by mild winters during
which most of the annual precipitation occurs (\Sexpr{18*24.5} to
\Sexpr{25*24.5} mm annually) and warm summers largely free of
precipitation events. Soils at the RES are classified as
Esquon-Neerdobe clays with roughly 2\% soil organic matter
\citep{ssurgo} in the top 15cm.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. Newly developed entries are typcically
evaluated in these trials for 3 to 5 years prior to
their official release. For these cultivars planted in breeding trials
at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
3--4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars and
included every publicly released medium grain cultivar developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release was
determined.

Experimental plots at the RES are managed similar to University of California
Cooperative Extension following prescribed best management
practice for land preparation, fertility rates, and pest
and disease control \citep{ucce}. As is typical for water-seeded
systems, for planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\tsups{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

To test the hypothesis, the influence of breeding and yield erosion
were quantified on yield over time using a Bayesian hierarchical
model. Yield was modeled as a response to the number of years since
the cultivar was officially released and the year of release using a
mixed-effects linear regression. To account for similarities between
years and between cultivars, year and cultivar were included in the
model as random effects.

@@

Data were processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the `rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package \citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval \citep{rstanarm}. To verify the estimates, the analysis was
also conducted using classical methods (frequentist mixed-effects
model fit by maximum likelihood; See supplemental material for
details). Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

@@

There was agreement between the estimates from the Bayesian and
classical approaches. The Bayesian estimates are presented here,
though the classical estimates can be found in the supplemental
material. Comparing the cultivars' yield during each's initial time
(average 3 years) in the experimental trials confirmed that genetic
yield improvements have been consistant over the 30 year study period
(Fig. \ref{fig:first_yrs}). As estimated by the model including both
yield erosion and yield improvement over time, there is no conclusive
evidence that the yields of medium grain rice cultivars selected and
released by CCRRF's breeding program have, on average, increased over
the last 30 years (Table \ref{tbl:coef}). The full model which
corrected for both the year of release and yield erosion was estimated
to have better out-of-sample predictive accuracy compared to reduced
models containing only one predictor (See supplemental material for
further details). The estimated change in grain yield per year in the
CCRRF's breeding trials at the RES was estimated to be a negligible
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:first_yrs}).

@@

plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
     yaxt = "n",
     ylim = c(8, 12.5),
     xaxt = "n",
     xlim = c(1980,2020))
axis(1, at = seq(1980, 2020, by = 10))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000,
       lty = 2)
with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
                         y1 = mean - se))
text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
     labels = gsub("M", "M-", row.names(plot_data2)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
mtext("Adjusting for yield erosion", adj = 0)

polygon(x = c(yrs, rev(yrs)),
        y = c(ints2[1,],rev(ints2[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since release",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8.0, 12.5))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
       b = yrs_coef/1000)
polygon(x = c(yrs_in, rev(yrs_in)),
        y = c(ints[1,],rev(ints[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))
text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
mtext("Adjusting for yield erosion", adj = 0)

@@

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other advancements due to plant
improvement, including increased quality \citep{mckenzie1993} and
milling yield \citep{mckenzie1994}, have been made over the study
period and are important. Also, crop duration has been reduced. For
example, during this study period, M-202 and M-206 have been the
predominate cultivars grown in CA, and while their yield potentials
were similar at the time of release (Fig. 5), M-206 matures about 6
days earlier than M-202 (Table \ref{tbl:ids}). Shorter duration is very important
in CA due to an already short growing season and water limitations
(shorter duration cultivars would presumably require less irrigation
water).
